---
title: "Modeling Data"
author: "Manda Egbert, John Dames, Kristina Burgess, and Elijah Weis"
date: "2023-04-12"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

### Load and Import
This chunk loads multiple packages to aid in organizing and modeling our data. It also uses read.csv to import our cleaned data set. 
```{r}
library(tidyverse)
library(modelr)
library(cvTools)
library(glmnet)
install.packages("reshape2")
library(reshape2)
set.seed(123)

clean.cat <- read.csv("C:/Users/amand/OneDrive/Desktop/IA Research/clean.categories.csv")
clean.num <- read.csv("C:/Users/amand/OneDrive/Desktop/IA Research/clean.numeric.csv")
```
### Gender by generation
Gender was split into 3 sections. Female, male, and other.  Other consists of 2 non-binary and 1 questioning individuals. These were combined because very few responded with these and having them separate made it difficult to differentiate on the graph.

```{r}
print(levels(clean.cat$generation))

# Change the order of the levels for the 'generation' variable
clean.cat$generation <- factor(clean.cat$generation, 
                                levels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"))

# Check the new order of the levels for the 'generation' variable
print(levels(clean.cat$generation))
```

```{r}
ggplot(clean.cat, aes(x = generation, fill = gender)) +
  geom_bar(position = "stack") +
  labs(title = "Gender Distribution by Generation",
       x = "Generation", y = "Count") +
  scale_fill_manual(values = c("#0072B2", "#E69F00", "purple"))
```

```{r}
ggplot(clean.cat, aes(x = gender, fill = gender)) +
  geom_bar() +
  scale_fill_manual(values = c("#0072B2", "#E69F00", "purple")) +
  labs(title = "Gender Counts", x = "Gender", y = "Count")
```

### Budget distribution (see outliers)
```{r Budget distibution}
# Select only the numeric values of groc.budget
groc.budget_numeric <- as.numeric(clean.cat$groc.budget)

# Identify outliers using the boxplot method
outliers <- boxplot.stats(groc.budget_numeric)$out

# Create a histogram
hist(groc.budget_numeric, breaks = 20, main = "Histogram of groc.budget")

# Add a vertical line for each outlier
abline(v = outliers, col = "red", lwd = 2)
```
### Boxplot: Grocery budget by generation
Note that boomers has the smallest amount of variation
```{r Budget by generation}
ggplot(clean.cat, aes(x = generation, y = groc.budget_numeric)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "text", size = 4, color = "black", 
               position = position_dodge(width = 0.75), show.legend = FALSE, 
               aes(label = round(..y.., 2))) +
  labs(title = "Boxplot of groc.budget for each category of generation",
       x = "Generation", y = "groc.budget")
```


### Grocery store by generation
```{r}
total_choices <- aggregate(choice ~ generation + store, data_long, sum)

# Calculate the proportion of each choice in each generation and store combination
prop_data <- total_choices %>%
  group_by(generation) %>%
  mutate(prop_choice = choice / sum(choice)) %>%
  ungroup() %>%
  select(-choice)

# Plot the proportion of each choice by generation, with bars stacked proportionally and labels for each section
ggplot(prop_data, aes(x = generation, y = prop_choice, fill = store, label = scales::percent(prop_choice))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Store Choices by Generation",
       x = "Generation", y = "Proportion") +
  scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73", "#CC79A7", "#D55E00", "purple")) +
  theme(legend.position = "bottom") +
  geom_text(position = position_fill(vjust = 0.5), size = 3, color = "white")
```

```{r}
# Load the dplyr and scales packages
library(dplyr)
library(scales)

# Group the data by generation and each store variable, and calculate the mean of each store variable
prop_data <- clean.cat %>%
  group_by(generation) %>%
  summarize(`Supermarket` = mean(supermarket),
            `Grocery Store` = mean(grocery.store),
            `Farmers Market` = mean(farmer.market),
            `Online Store` = mean(online.groc),
            `Warehouse` = mean(warehouse),
            `Online Prepping Service` = mean(online.prep))

# Create a new data frame with the proportions converted to percentages
prop_data_percent <- prop_data %>%
  mutate(across(-generation, percent))

# Show the proportion of store choices by generation as percentages
prop_data_percent
```



### PErception vs reality
```{r Millenial}
# Calculate mean perception rating for each generation
perception_data <- aggregate(impact.boomer ~ generation, clean.cat, mean)

# Calculate mean self rating for millennial
self_data <- aggregate(impact ~ generation, clean.cat, mean)
self_data$generation <- "Baby Boomer"

# Merge perception and self data using 'generation' as the key
combined_data <- merge(perception_data, self_data, by = "generation", all = TRUE)

# Melt data into long format
combined_data_long <- tidyr::gather(combined_data, variable, rating, -generation, na.rm = TRUE)

# Set the order of the levels for the 'generation' variable
combined_data_long$generation <- factor(combined_data_long$generation, 
                                        levels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"))

# Plot grouped bar graph
ggplot(combined_data_long, aes(x = generation, y = rating, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Perceived vs. Personal Effects of Increasing Grocery Prices on Baby",
       x = "Generation", y = "Rating") +
  scale_fill_manual(values = c("#E69F00", "#0072B2")) +
  theme(legend.position = "bottom")
```
```{r Gen Z}
perception_data <- aggregate(impact.genZ ~ generation, clean.cat, mean)

self_data <- aggregate(impact ~ generation, clean.cat, mean)
self_data$generation <- "Gen Z"

combined_data <- merge(perception_data, self_data, by = "generation", all = TRUE)

combined_data_long <- tidyr::gather(combined_data, variable, rating, -generation, na.rm = TRUE)

ggplot(combined_data_long, aes(x = generation, y = rating, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Perceived vs. Actual Effects of Increasing Grocery Prices on Gen Z",
       x = "Generation", y = "Rating") +
  scale_fill_manual(values = c("#E69F00", "#0072B2")) +
  theme(legend.position = "bottom")
```
```{r Boomer}
perception_data <- aggregate(impact.boomer ~ generation, clean.cat, mean)

self_data <- aggregate(impact ~ generation, clean.cat, mean)
self_data$generation <- "Baby Boomers"

combined_data <- merge(perception_data, self_data, by = "generation", all = TRUE)

combined_data_long <- tidyr::gather(combined_data, variable, rating, -generation, na.rm = TRUE)

ggplot(combined_data_long, aes(x = generation, y = rating, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Perceived vs. Actual Effects of Increasing Grocery Prices on Baby Boomers",
       x = "Generation", y = "Rating") +
  scale_fill_manual(values = c("#E69F00", "#0072B2")) +
  theme(legend.position = "bottom")
```
```{r Gen X}
perception_data <- aggregate(impact.genX ~ generation, clean.cat, mean)

self_data <- aggregate(impact ~ generation, clean.cat, mean)
self_data$generation <- "Gen X"

combined_data <- merge(perception_data, self_data, by = "generation", all = TRUE)

combined_data_long <- tidyr::gather(combined_data, variable, rating, -generation, na.rm = TRUE)

ggplot(combined_data_long, aes(x = generation, y = rating, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Perceived vs. Actual Effects of Increasing Grocery Prices on Gen X",
       x = "Generation", y = "Rating") +
  scale_fill_manual(values = c("#E69F00", "#0072B2")) +
  theme(legend.position = "bottom")
```

### Simple Linear Regression
The following chunks create a linear model using a single explanatory variable,  to predict our response variable, Results.  We will use the output to consider whether the variable is significant (p-value less than 5%) and whether r-squared is high enough (preferably greater than 80%, but we will be satisfied with just the highest value we can find) to suggest the regression does a good job at explaining the variation in our response.

Insigificant, however note that boomer is the best predictor and this goes well with the limited variation we see in the boxplot. millenial is kinda sig
```{r Budget}
lm.budget <- lm(generation ~ groc.budget_numeric, data=clean.num)
summary(lm.budget)
```

Good pvalue, bad r-sq
```{r Household}
lm.household <- lm(generation ~ household.size, data=clean.num)
summary(lm.household)
```
Look into whether millennial are more likely to shop at online grocery stores AND online prep
```{r Online grocery shopping}
lm.online.groc <- lm(generation ~ online.groc, data=clean.num)
summary(lm.online.groc)
```


### Multiple Linear Regression
The following chunks create a multiple linear model using multiple explanatory variables to predict our response variable, Results.  We will use the output to determine which variables are significant (p-value less than 5%) and whether r-squared is high enough (preferably greater than 80%) to suggest the regression does a good job at explaining the variation in our response.

```{r Full model}
model.all <- lm(generation ~ groc.budget_numeric + household.size + online.groc + income + concern + changes1, data=clean.num)
summary(model.all)
```

```{r}
clean.cat_long <- tidyr::pivot_longer(clean.cat, cols = c("price", "quality", "brand", "convenience"), 
                                      names_to = "factor", values_to = "value")

# Calculate the total number of selections for each factor and generation combination
total_selections <- aggregate(value ~ generation + factor, clean.cat_long, sum)

# Calculate the proportion of each factor selected by generation
prop_data <- total_selections %>%
  group_by(generation) %>%
  mutate(prop_factor = value / sum(value)) %>%
  ungroup() %>%
  select(-value)

# Convert proportions to percentages and create a new column in the prop_data dataframe
prop_data_percent <- prop_data %>%
  mutate(prop_percent = scales::percent(prop_factor))

# Create a stacked bar chart for each generation with proportions displayed in labels
ggplot(prop_data_percent, aes(x = generation, y = prop_factor, fill = factor, label = prop_percent)) +
  geom_col(position = "fill") +
  labs(title = "Factors Most Important to Each Generation",
       x = "Generation", y = "Proportion", fill = "Factor") +
  scale_fill_manual(values = c("#0072B2", "#E69F00", "#009E73", "#CC79A7")) +
  theme(legend.position = "bottom") +
  geom_text(position = position_fill(vjust = 0.5), color = "white")
```



```{r}
model.reduced <- lm(generation ~ household.size + changes1, data=clean.num)
summary(model.reduced)
```


### Chi-Sq
Insig

```{r}
# create a subset of the data with only the before strategy variables and the generation variable
before_data <- clean.num[, c("generation", "before1", "before2", "before3", "before4", "before5", "before6", "before7", "before8", "before9", "before10", "before11", "before12", "before13")]

# run the chi-squared test
before_chi <- chisq.test(before_data[, -1])
before_chi
```
```{r}
# create a subset of the data with only the after strategy variables and the generation variable
after_data <- clean.num[, c("generation", "after1", "after2", "after3", "after4", "after5", "after6", "after7", "after8", "after9", "after10", "after11", "after12", "after13")]

# run the chi-squared test
after_chi <- chisq.test(after_data[, -1])
after_chi
```

### Stacked bar-graph: Generation vs Strategies
```{r}
df <- data.frame(
  generation = clean.cat$generation,
  after1 = clean.cat$after1,
  after2 = clean.cat$after2,
  after3 = clean.cat$after3,
  after4 = clean.cat$after4,
  after5 = clean.cat$after5,
  after6 = clean.cat$after6,
  after7 = clean.cat$after7,
  after8 = clean.cat$after8,
  after9 = clean.cat$after9,
  after10 = clean.cat$after10,
  after11 = clean.cat$after11,
  after12 = clean.cat$after12,
  after13 = clean.cat$after13
)

df.counts <- aggregate(df[,-1], by = list(df$generation), FUN = sum)

# Calculate the proportion of respondents in each generation who used each money-saving strategy
df.prop <- df.counts
df.prop[, -1] <- df.counts[, -1] / rowSums(df.counts[, -1])

# Reshape the data to long format for plotting
df.long <- reshape2::melt(df.prop, id.vars = "Group.1")

# Create a stacked bar chart using ggplot2
ggplot(data = df.long, aes(x = Group.1, y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  xlab("Generation") +
  ylab("Proportion of respondents") +
  ggtitle("Money-saving strategies used after the increase in grocery prices") +
  scale_fill_manual(values = c("#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17BECF", "#9F80A0", "#CCBEA0", "#DB8A1B"))
```
```{r}
df <- data.frame(
  generation = clean.cat$generation,
  before1 = clean.cat$before1,
  before2 = clean.cat$before2,
  before3 = clean.cat$before3,
  before4 = clean.cat$before4,
  before5 = clean.cat$before5,
  before6 = clean.cat$before6,
  before7 = clean.cat$before7,
  before8 = clean.cat$before8,
  before9 = clean.cat$before9,
  before10 = clean.cat$before10,
  before11 = clean.cat$before11,
  before12 = clean.cat$before12,
  before13 = clean.cat$before13
)

df.counts <- aggregate(df[,-1], by = list(df$generation), FUN = sum)

# Calculate the proportion of respondents in each generation who used each money-saving strategy
df.prop <- df.counts
df.prop[, -1] <- df.counts[, -1] / rowSums(df.counts[, -1])

# Reshape the data to long format for plotting
df.long <- reshape2::melt(df.prop, id.vars = "Group.1")

# Create a stacked bar chart using ggplot2
ggplot(data = df.long, aes(x = Group.1, y = value, fill = variable)) +
  geom_bar(stat = "identity") +
  xlab("Generation") +
  ylab("Proportion of respondents") +
  ggtitle("Money-saving strategies used before the increase in grocery prices") +
  scale_fill_manual(values = c("#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B", "#E377C2", "#7F7F7F", "#BCBD22", "#17BECF", "#9F80A0", "#CCBEA0", "#DB8A1B"))
```


```{r 1}
# Define the maximum count across all subsets of the data
max_count <- max(table(clean.cat$generation))

# Create a subset of the data for respondents who selected 1 in the "after1" variable
after11_yes <- subset(clean.cat, after11 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after11_yes$generation)

# Define custom colors for the bars
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a vertical bar graph with custom colors and labels
barplot(generation_counts, horiz = FALSE, col = colors, main = "Generational Switch to Buying Closer to Expiration", xlab = "Generation", ylab = "Number of Respondents", names.arg = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), ylim = c(0, max_count))
```

INFLATION
38% gen x
30% baby boomer
boomer mentions biden and supply chain shortages
gen x mentions biden A LOT
gen z mostly inflation 

```{r 2}
df_sub <- clean.cat[, c("before2", "after2", "generation")]

library(dplyr)
df_summary <- df_sub %>%
  group_by(generation) %>%
  summarize(before2_count = sum(before2 == 1),
            after2_count = sum(after2 == 1),
            before2_prop = before2_count / n(),
            after2_prop = after2_count / n())

# Reshape the data from wide to long format
df_long <- pivot_longer(df_summary, 
                        cols = c("before2_prop", "after2_prop"), 
                        names_to = "variable", 
                        values_to = "value")

# Create the bar plot
ggplot(df_long, aes(x = generation, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("blue", "red"), name = "",
                    labels = c("Before", "After")) +
  labs(x = "Generation", y = "Proportion",
       title = "Generational Switch to Coupons and Discounts")
```


```{r 4}
df_sub <- clean.cat[, c("before4", "after4", "generation")]

library(dplyr)
df_summary <- df_sub %>%
  group_by(generation) %>%
  summarize(before4_count = sum(before4 == 1),
            after4_count = sum(after4 == 1),
            before4_prop = before4_count / n(),
            after4_prop = after4_count / n())

# Reshape the data from wide to long format
df_long <- pivot_longer(df_summary, 
                        cols = c("before4_prop", "after4_prop"), 
                        names_to = "variable", 
                        values_to = "value")

# Create the bar plot
ggplot(df_long, aes(x = generation, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("blue", "red"), name = "",
                    labels = c("Before", "After")) +
  labs(x = "Generation", y = "Proportion",
       title = "Generational Switch to Comparison Shopping")

```

### Pie Charts

```{r 1}
after1_yes <- subset(clean.cat, after1 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after1_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to More Affordable Grocery Store
", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```
```{r 2}
after2_yes <- subset(clean.cat, after2 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after2_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Using Coupons or Discounts More Frequently", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 3}
after3_yes <- subset(clean.cat, after3 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after3_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Reducing the Number of Items Purchased", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 4}
after4_yes <- subset(clean.cat, after4 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after4_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Seeking Out Lower-Cost or Generic Brands", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 5}
after5_yes <- subset(clean.cat, after5 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after5_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Shopping at Several Stores to Find Deals", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 6}
after6_yes <- subset(clean.cat, after6 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after6_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to using Technology to Find Deals", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 7}
after7_yes <- subset(clean.cat, after7 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after7_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Avoiding Snacks or Non-Meal Food Items", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 8}
after8_yes <- subset(clean.cat, after8 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after8_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Avoiding Store When Hungry
", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 9}
after9_yes <- subset(clean.cat, after9 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after9_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Buying in Bulk", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 10}
after10_yes <- subset(clean.cat, after10 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after10_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Favoring Items with a Long Shelf Life", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```

```{r 11}
after11_yes <- subset(clean.cat, after11 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after11_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Generational Switch to Buying Closer to Expiration for a Low Price", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```


other:
```{r 12}
after12_yes <- subset(clean.cat, after12 == 1)

# Create a table of counts of respondents by generation
generation_counts <- table(after12_yes$generation)

# Create a vector of custom colors
colors <- c("#F44336", "#FFC107", "#9C27B0", "#2196F3")

# Create a pie chart with custom colors
pie(generation_counts, main = "Proportion of Respondents Who Selected 1 in the 'after12' Variable\n by Generation", labels = c("Baby Boomers", "Gen X", "Millennial", "Gen Z"), col = colors)
```




